---
title: "DATA 607 Week 3B - Window Functions"
author: "Patricio Romero"
date: "2026-09-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

SQL window functions perform calculations across groups of related rows without reducing the dataset to one row per group. They are useful for analyzing time-series data because each original observation can remain visible while cumulative and moving statistics are calculated.

This project uses daily stock-price data for Apple, Microsoft, and Google. PostgreSQL stores the observations, and SQL window functions are used to calculate a year-to-date average and a six-day moving average for each company.

## Data Collection

Daily adjusted closing prices were retrieved from Yahoo Finance using the R package `quantmod` and its `getSymbols()` function.

The dataset contains daily adjusted closing prices for the following companies:

- Apple (`AAPL`)
- Microsoft (`MSFT`)
- Alphabet/Google (`GOOGL`)

The observations begin on January 3, 2022, and include one row for each available trading day and company. The latest observations used in this approach submission are dated September 16, 2026.

Each record contains:

- `price_date`: trading date
- `symbol`: stock ticker
- `company`: company name
- `adjusted_close`: adjusted closing price

Non-trading days were not created or estimated. Only available trading-day observations were stored.

## Planned Approach

The project will follow these steps:

1. Obtain daily adjusted closing prices for the three selected companies.
2. Combine the observations into a consistent long-format dataset.
3. Validate dates, company identifiers, prices, and duplicate records.
4. Import the prepared data into PostgreSQL.
5. Confirm the number and date range of observations for each company.
6. Use a SQL window function to calculate the year-to-date average for each company and calendar year.
7. Use a second SQL window function to calculate the six-day moving average for each company.
8. Preserve each daily observation while adding the calculated averages.
9. Review representative dates to confirm that the window calculations are correct.
10. Present the SQL code, results, and conclusions in the final report.

The year-to-date calculation will partition the observations by company and calendar year, order them by date, and include all available records from the beginning of the year through the current row.

The six-day moving calculation will partition the observations by company, order them by date, and use the current observation together with the five preceding trading-day observations.

## Database Design

The observations will be stored in a PostgreSQL table named `daily_prices`.

The table will contain:

- `price_id`: generated primary key
- `price_date`: trading date
- `symbol`: stock ticker
- `company`: company name
- `adjusted_close`: adjusted closing price

A uniqueness rule will prevent more than one observation for the same stock symbol and trading date.

PostgreSQL and pgAdmin 4 will be used to store, inspect, validate, and later analyze the dataset. The calculations will not be performed until the data have been imported and validated.

## Data Retrieval from PostgreSQL

The prepared observations were imported into the PostgreSQL table `daily_prices`. R retrieves the stored data to confirm that the database contains the expected records before any window calculations are performed.

```{r load-database-packages, message=FALSE, warning=FALSE}
library(DBI)
library(RPostgres)
```

```{r database-connection, echo=FALSE, message=FALSE, warning=FALSE}
con <- dbConnect(
  RPostgres::Postgres(),
  host = "127.0.0.1",
  port = 55433,
  dbname = "data607_week3",
  user = "postgres",
  password = Sys.getenv("DATA607_DB_PASSWORD")
)
```

```{r retrieve-data}
daily_prices <- dbGetQuery(
  con,
  "
  SELECT price_date, symbol, company, adjusted_close
  FROM daily_prices
  ORDER BY symbol, price_date
  "
)

head(daily_prices)
```

## Data Validation

The imported data were validated before applying any SQL window functions.

```{r validate-data}
data.frame(
  total_observations = nrow(daily_prices),
  companies = length(unique(daily_prices$symbol)),
  missing_dates = sum(is.na(daily_prices$price_date)),
  missing_symbols = sum(is.na(daily_prices$symbol) | daily_prices$symbol == ""),
  missing_companies = sum(is.na(daily_prices$company) | daily_prices$company == ""),
  duplicate_symbol_dates = sum(
    duplicated(daily_prices[c("symbol", "price_date")])
  ),
  missing_prices = sum(is.na(daily_prices$adjusted_close))
)
```

## Observation Summary by Company

The number of observations and the available date range were confirmed for each stock symbol before applying the window calculations.

```{r observation-summary}
dbGetQuery(
  con,
  "
  SELECT
      symbol,
      company,
      COUNT(*) AS observations,
      MIN(price_date) AS first_date,
      MAX(price_date) AS last_date
  FROM daily_prices
  GROUP BY symbol, company
  ORDER BY symbol;
  "
)
```

## Year-to-Date Average

The year-to-date average is calculated independently for each stock symbol and calendar year. The window begins with the first available trading day of the year and continues through the current observation.

```{r ytd-average}
ytd_prices <- dbGetQuery(
  con,
  "
  SELECT
      price_date,
      symbol,
      company,
      adjusted_close,
      AVG(adjusted_close) OVER (
          PARTITION BY symbol, DATE_PART('year', price_date)
          ORDER BY price_date
          ROWS BETWEEN UNBOUNDED PRECEDING AND CURRENT ROW
      ) AS ytd_average
  FROM daily_prices
  ORDER BY symbol, price_date;
  "
)

head(ytd_prices)
```
## Six-Day Moving Average

The six-day moving average is calculated independently for each stock symbol. The window includes the current trading-day observation and the five preceding trading-day observations.

```{r six-day-moving-average}
moving_prices <- dbGetQuery(
  con,
  "
  SELECT
      price_date,
      symbol,
      company,
      adjusted_close,
      AVG(adjusted_close) OVER (
          PARTITION BY symbol
          ORDER BY price_date
          ROWS BETWEEN 5 PRECEDING AND CURRENT ROW
      ) AS six_day_moving_average
  FROM daily_prices
  ORDER BY symbol, price_date;
  "
)

head(moving_prices, 10)
```

## Combined Window Functions

The two window calculations can be included in the same SQL query. This preserves every original daily stock-price observation while adding both the year-to-date average and the six-day moving average.

```{r combined-window-functions}
window_results <- dbGetQuery(
  con,
  "
  SELECT
      price_date,
      symbol,
      company,
      adjusted_close,

      AVG(adjusted_close) OVER (
          PARTITION BY symbol, DATE_PART('year', price_date)
          ORDER BY price_date
          ROWS BETWEEN UNBOUNDED PRECEDING AND CURRENT ROW
      ) AS ytd_average,

      AVG(adjusted_close) OVER (
          PARTITION BY symbol
          ORDER BY price_date
          ROWS BETWEEN 5 PRECEDING AND CURRENT ROW
      ) AS six_day_moving_average

  FROM daily_prices
  ORDER BY symbol, price_date;
  "
)

head(window_results, 10)
```

## Representative Date Verification

Representative observations were reviewed for each stock symbol to confirm that the year-to-date and six-day moving averages were calculated as expected.

```{r representative-dates}
representative_results <- window_results[
  window_results$symbol %in% c("AAPL", "GOOGL", "MSFT") &
  window_results$price_date %in% as.Date(c(
    "2022-01-03",
    "2022-01-10",
    "2022-02-01"
  )),
]

representative_results
```


## Window Results Validation

The window-function results are validated by comparing the number of calculated observations with the original dataset and checking for missing calculated values.

```{r validate-window-results}
data.frame(
  original_observations = nrow(daily_prices),
  calculated_observations = nrow(window_results),
  missing_ytd_values = sum(is.na(window_results$ytd_average)),
  missing_moving_values = sum(is.na(window_results$six_day_moving_average))
)
```

## Results and Conclusion

The implementation followed the planned approach. The stock-price observations were stored in PostgreSQL and analyzed using SQL window functions.

The year-to-date average was calculated independently for each stock symbol and calendar year. For each observation, the window included all available trading-day records from the beginning of the year through the current row.

The six-day moving average was calculated independently for each stock symbol using the current trading-day observation and the five preceding trading-day observations. For the first observations of each stock, PostgreSQL used the available rows until a complete six-observation window was available.

The window functions preserved the original daily stock-price observations while adding the calculated averages to each row. The validation results confirmed that the number of calculated observations matched the number of original observations and that the calculated window values contained no missing values.

This analysis demonstrates how SQL window functions can be used to calculate cumulative and moving statistics without reducing the original dataset through aggregation.

## AI Use

ChatGPT was used to help interpret the assignment requirements, organize the planned approach, improve the English writing, select an appropriate database structure, and provide coding guidance. I collected and imported the data, ran the validation code, reviewed the results, and confirmed the conclusions myself.
