---
title: "DATA 607 Week 3B - Window Functions"
author: "Patricio Romero"
date: "2026-09-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

SQL window functions perform calculations across groups of related rows without reducing the dataset to one row per group. They are useful for analyzing time-series data because each original observation can remain visible while cumulative and moving statistics are calculated.

This project will use daily stock-price data for Apple, Microsoft, and Google. PostgreSQL will store the observations, and SQL window functions will later calculate a year-to-date average and a six-day moving average for each company.

## Data Collection

The dataset will contain daily adjusted closing prices for the following companies:

- Apple (`AAPL`)
- Microsoft (`MSFT`)
- Alphabet/Google (`GOOGL`)

The observations will begin in January 2022 and will include one row for each available trading day and company.

Each record will contain:

- `price_date`: trading date
- `symbol`: stock ticker
- `company`: company name
- `adjusted_close`: adjusted closing price

Non-trading days will not be created or estimated. Only available trading-day observations will be stored.

## Planned Approach

The project will follow these steps:

1. Obtain daily adjusted closing prices for the three selected companies.
2. Combine the observations into a consistent long-format dataset.
3. Validate dates, company identifiers, prices, and duplicate records.
4. Import the prepared data into PostgreSQL.
5. Confirm the number and date range of observations for each company.
6. Use a SQL window function to calculate the year-to-date average for each company and calendar year.
7. Use a second SQL window function to calculate the six-day moving average for each company.
8. Preserve each daily observation while adding the calculated averages.
9. Review representative dates to confirm that the window calculations are correct.
10. Present the SQL code, results, and conclusions in the final report.

The year-to-date calculation will partition the observations by company and calendar year, order them by date, and include all available records from the beginning of the year through the current row.

The six-day moving calculation will partition the observations by company, order them by date, and use the current observation together with the five preceding trading-day observations.

## Database Design

The observations will be stored in a PostgreSQL table named `daily_prices`.

The table will contain:

- `price_id`: generated primary key
- `price_date`: trading date
- `symbol`: stock ticker
- `company`: company name
- `adjusted_close`: adjusted closing price

A uniqueness rule will prevent more than one observation for the same stock symbol and trading date.

PostgreSQL and pgAdmin 4 will be used to store, inspect, validate, and later analyze the dataset. The calculations will not be performed until the data have been imported and validated.

## Data Retrieval from PostgreSQL

The prepared observations were imported into the PostgreSQL table `daily_prices`. R retrieves the stored data to confirm that the database contains the expected records before any window calculations are performed.

```{r load-database-packages, message=FALSE, warning=FALSE}
library(DBI)
library(RPostgres)
```

```{r database-connection, echo=FALSE, message=FALSE, warning=FALSE}
con <- dbConnect(
  RPostgres::Postgres(),
  host = "127.0.0.1",
  port = 55433,
  dbname = "data607_week3",
  user = "postgres",
  password = Sys.getenv("DATA607_DB_PASSWORD")
)
```

```{r retrieve-data}
daily_prices <- dbGetQuery(
  con,
  "
  SELECT price_date, symbol, company, adjusted_close
  FROM daily_prices
  ORDER BY symbol, price_date
  "
)

head(daily_prices)
```

## Data Validation

The imported data were validated before applying any SQL window functions.

```{r validate-data}
data.frame(
  total_observations = nrow(daily_prices),
  companies = length(unique(daily_prices$symbol)),
  duplicate_symbol_dates = sum(
    duplicated(daily_prices[c("symbol", "price_date")])
  ),
  missing_prices = sum(is.na(daily_prices$adjusted_close))
)
```


## AI Use

ChatGPT was used to help interpret the assignment requirements, organize the planned approach, improve the English writing, select an appropriate database structure, and provide coding guidance. I collected and imported the data, ran the validation code, reviewed the results, and confirmed the conclusions myself.